---
title: "Finding Outliers"
output: 
  html_document:
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.align = "center", 
  collapse = TRUE,
  comment = "#>"
)
```


```{r setup, message = FALSE}
library(seahorse)
library(tidyverse)
library(scales)
```

## Introduction

One of the main goals of the `seahorse` package is to facilitate outlier 
detection both among technical replicates in a `Seahorse` object and among 
biological replicates across experiments in a `Herd` object. This document 
serves to illustrate the approach taken by this package and to record the 
rationale for decisions.

## Data

First, we will create a `seahorse` object. Since this automatically identifies 
outliers, we will remove all blank and outlier assignments prior to our 
analysis. 

```{r, message = FALSE}
path <- 
  system.file(
    "extdata/raw_1.xlsx", 
    package = "seahorse", 
    mustWork = TRUE
  )

cells <- 
  format_cells(
    system.file(
      "extdata/counts_1.csv", 
      package = "seahorse", 
      mustWork = TRUE
    )
  )

x <- 
  Seahorse(
    path = path, 
    wells = wells_ex, 
    stages = stages_mst,
    cells = cells
  )
blanks(x, "remove") <- NA
outliers(x, "remove") <- NA

df <- rates(x, blanks = TRUE, outliers = TRUE, normalize = FALSE)
```

```{r}
p1 <- 
  ggplot(df) +
  facet_grid(
    rows = vars(rate), 
    cols = vars(group), 
    scales = "free_y"
  ) +
  aes(
    x = measurement, 
    y = value
  ) +
  geom_line(
    aes(
      color = group, 
      group = well
    ), 
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = pretty_breaks())
p1 +
  labs(title = "Raw data")
```

## Identifying Outliers

The data for each group are fit using a robust linear model accounting for time 
and stage. Several model formulas were considered. The current formula, _value ~ 
measurement * stage_, accounts for changing data values within an interval 
across subsequent measurements. An alternative approach uses _value ~ 
factor(measurement)_, which will fit the median of each measurement as the 
summary values. Generally, robust linear models with `MASS::rlm` returned 
similar values as ordinary least square fits with `stats::lm`. 

```{r}
models <- 
  df |> 
  group_by(rate, group, type) |> 
  nest() |> 
  mutate(
    model = map(
      data, 
      \(x) MASS::rlm(value ~ measurement * stage, data = x, maxit = 100)
    ), 
    fit = map(model, fitted),
    res = map(model, residuals),
    se = map(res, \(x) x ^ 2)
  ) |> 
  unnest(c(data, fit, res, se))

fits <- 
  models |> 
  select(type:stage, fit)
```

```{r}
p1 + 
  geom_line(
    data = fits, 
    aes(y = fit), 
    linewidth = 1
  ) +
  labs(
    title = "Model fits"
  )
```

The squared residuals between the model fit and data values for each 
well are calculated. The wells with a sum of squared residuals greater than 
four times the median absolute deviation for the group are identified as 
outliers.

```{r, message = FALSE, tab.cap = "Groups with outliers", rows.print = 5}
out <- 
  models |> 
  group_by(rate, group, well) |> 
  summarize(sse = sum(se)) |> 
  mutate(outlier = abs(sse - stats::median(sse)) / stats::mad(sse) > 4) 

out |> 
  group_by(rate, group) |> 
  filter(any(outlier))
```

```{r}
df |> 
  left_join(out, by = c("rate", "group", "well")) |> 
  ggplot() +
  facet_grid(
    rows = vars(rate), 
    cols = vars(group), 
    scales = "free_y"
  ) +
  aes(
    x = measurement, 
    y = value
  ) +
  geom_line(
    aes(
      color = group, 
      group = well, 
      linetype = outlier
    ), 
    show.legend = FALSE
  ) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(
    title = "Outlier wells"
  )
```

## Notes

Using a similar approach, one can attempt to identify outlying data points. I 
was not able to find an approach that successfully excluded relatively few 
outlying data points. 

One group has suggested that these analysis should use log-transformed rate 
values. In these example data, better fits were acheived with non-transformed 
values. 

Outlier wells should be identified after normalization. In this example, 
non-normalized values were analyzed for aesthetic reasons related to blank 
value plotting.
